---
title: "Contributor Guidelines"
description: "This page outlines the development process for `{gsm}` packages, including how to contribute by filing issues, bug reports, and submitting code via a Pull Request."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contributor Guidelines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(gsm.core)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This page outlines the development process for `{gsm}` packages. It summarizes both our project management and development workflows. The project management workflow focuses on using issues to are capture user requirements, bugs and technical requirements. The development workflow explains how to submit code using the GitHub flow paradigm.


# Project Management

The project management workflow focuses on using issues to capture user requirements, bugs, and technical requirements. We also use the [gsm Roadmap](https://github.com/orgs/Gilead-BioStats/projects/41) GitHub Project to track issues and set priorities across repositories. 


## Issues

Issues are the primary way to communicate what needs to be done and to track progress. Several issue templates are provided to help streamline this process, including: 

- [Requirements](https://github.com/Gilead-BioStats/gsm.core/issues/new?template=business_requirement.md): Use this template to create a User Requirement. Requirements are then assigned sub-issues using the issue types below. 
- [Bug](https://github.com/Gilead-BioStats/gsm.core/issues/new?template=bug_report.md): Bug reports for when something isn't working
- [Feature](https://github.com/Gilead-BioStats/gsm.core/issues/new?template=feature_request.md): User-facing Functionality
- [Technical Task](https://github.com/Gilead-BioStats/gsm.core/issues/new?template=qc_update.md): Non-user facing tasks, such as infrastructure updates or internal tooling improvements.
- [Documentation Task](https://github.com/Gilead-BioStats/gsm.core/issues/new?template=documentation.md): Improvements or additions to the documentation including function docs, readme updates and vignettes.

The issue templates are maintained in {[gsm.utils](https://github.com/Gilead-BioStats/gsm.utils)} and automatically updated in all gsm packages whenever updates are made. Note that suggestions or other input that might not warrant formal submission of an issue can be filed under [discussions](https://github.com/Gilead-BioStats/gsm.core/discussions), which can help facilitate discourse of specific use-cases or requests.

---

## Road Map

The [gsm Roadmap](https://github.com/orgs/Gilead-BioStats/projects/41) GitHub Project is used to track issues and set priorities across projects. The gsm team generally batches work in to quarterly releases, although package releases can happen more frequently. Key cross-repo views in the project include: 

- Roadmap - a list of issues split out by quarter. Generally speaking, related functionality should be grouped in to a Requirement. Key columns include: 
   - Status - Issue status (e.g. To Do, In Progress, Done)
   - Repository/Milestone
   - Roadmap - Quarter where the requirement is planned for completion
   - Triage - has the requirement been approved for development work
   - Type - Bug, Feature, Task, Documentation
- Issues - a list of all open non-requirement issues, split by type.
- Releases - All open issues grouped by repo and sorted by milestone. 

# Development Workflow

Development is done using R package development best practices ([R Packages](https://r-pkgs.org/)) and the **GitHub Flow** workflow ([GitHub Flow Guide](https://docs.github.com/en/get-started/using-github/github-flow)).
The goal is to keep code reliable, traceable, and easy for everyone to contribute.

---

## Branches

We use a simple branching model to keep work organized:

* **`main`** – stable, production-ready code. Every package release comes from here.
* **`dev`** – working branch that collects all new features, bugfixes, and QC tasks before release.
* **`fix-*`** – short-lived branches for specific issues (feature, bug, or QC task). All development starts here.
* **`release-*`** – temporary branches for regression testing, QC, and documentation before merging into `main`.

👉 This structure ensures that unfinished work never goes directly into production, but can still be tested and reviewed in isolation.

---

## Development Workflow

1. **Open an issue** – every change begins with an issue using the right template (`Bug`, `Feature`,`Technical Task`, or `Documentation Task`). This creates a clear record of what’s being worked on.
2. **Create a branch** – branch names link to issues (e.g., `fix-111` for issue #111) so progress is traceable.
3. **Develop & document** – write clean, consistent code, add roxygen2 documentation, and create/update [testthat](https://testthat.r-lib.org/) unit tests.
4. **Open a Pull Request** – submit your branch to `dev`, assign yourself, request a reviewer, and link the issue. CI checks will run automatically.
5. **Merge & clean up** – once approved and passing checks, merge into `dev` and delete the `fix-*` branch.

👉 Following this flow ensures that every contribution is tracked, tested, and reviewed before becoming part of the development branch.

---

## Pull Requests

Pull Requests are where code review happens. To keep them smooth:

* Follow the [tidyverse style guide](https://style.tidyverse.org/) See the Code Style section below for details on styling code per GSM standards.
* All functions must have **roxygen2 docs** (`@param`, `@return`, examples, etc.).
* All changes must include **unit tests** showing expected behavior.
* If Qualification double-programming tests were required per the content of the issue(s) tied to the PR, these tests must exist and be passing.
* Always **link PRs to their issues** using [closing keywords](https://docs.github.com/en/issues/tracking-your-work-with-issues/linking-a-pull-request-to-an-issue-using-a-keyword).
* PRs must pass all **CI checks** (tests, style, docs, coverage) before merge.
* Default target branch is `dev`; only **release** PRs target `main`.

👉 A well-prepared PR makes reviewing faster and ensures that releases are low-risk.

---

## Release Workflow

Releases move code from development into production. The goal is to ensure **reliability, traceability, and reproducibility**:

1. **Create a `release-x.y.z` branch** from `dev` (using [semantic versioning](https://semver.org/)).
2. **Prepare the release** – update version, NEWS, documentation, qualification files, and run checks (`devtools::check()`, spell check, `pkgdown::build_site()`).
3. **Open a release PR** into `main` using the release template. Assign yourself and request QC review.
4. **QC reviewers** complete checklists and may request fixes. Any fixes are made in `fix-*` branches and merged back into the release branch.
5. **Complete the release** – once approved, follow the following steps:
    * Merge PR into `main`
    * Create a GitHub Release
    * Attach QC reports
    * Create a PR that syncs `main` back into `dev`

👉 This process makes sure that every release is tested, documented, and signed off before it goes live.

---

## Code Style

We follow the [tidyverse style guide](https://style.tidyverse.org/) with minor tweaks.
Before each release, code is standardized with `styler` so formatting is consistent across the whole repo.

```{r eval = FALSE}
double_indent_style <- styler::tidyverse_style()
double_indent_style$indention$unindent_fun_dec <- NULL
double_indent_style$indention$update_indention_ref_fun_dec <- NULL
double_indent_style$line_break$remove_line_breaks_in_fun_dec <- NULL
styler::style_dir("R", transformers = double_indent_style)
styler::style_dir("tests", recursive = TRUE, transformers = double_indent_style)
```

---

## Continuous Integration

GitHub Actions run automatically on all Pull Requests to `dev` and `main` to catch issues early:

* **R CMD check** – confirms the package builds and passes tests.
* **Pkgdown** – Generates `pkgdown` docs.
* **Release checks** – run on multiple platforms (Linux, macOS, Windows) to confirm reproducibility.
* **Qualification Reports** – built and attached to releases for traceability. *Only run on merges to `main`.*

👉 Contributors don’t typically need to configure anything—just push code and the checks will run.

---

# Appendix – Quick Reference

### Fix Branch Workflow

1. Open issue with correct template.
2. Create `fix-*` branch.
3. Develop code + docs + tests.
4. Open PR to `dev` (assign self, request review, link issue).
5. Merge after approval + passing checks.

### Release Branch Workflow

1. Create `release-x.y.z` branch from `dev`.
2. Update version, NEWS, docs, qualification files.
3. Run checks locally + CI.
4. Open release PR to `main` with template.
5. QC reviewers complete checklist.
6. Merge, publish GitHub Release, attach QC report.
7. Sync `main` → `dev`, close issues, clean up branches.

### QC Checklist

* [ ] Roxygen2 docs complete (`@export`, `@param`, `@return`, examples).
* [ ] Unit tests for inputs, outputs, error handling.
* [ ] `devtools::check()` passes (no errors/warnings/notes).
* [ ] No sensitive data or hardcoded paths.
* [ ] Qualification specs + reports updated.
* [ ] All GitHub Actions checks pass.
